Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-09-11 Documents/M1/realizableFermionicM1/manuscript.tex
Moved Ran to first author.  Minor edits to abstract from Cory

%% Group authors per affiliation:
\author[ornl,utk-phys,jics]{Eirik Endeve\corref{cor}}
\ead{endevee@ornl.gov}

\author[utk-phys]{Ran Chu}
\ead{rchu@vols.utk.edu}

@@

\begin{abstract}
Building on the framework of Zhang \& Shu \cite{zhangShu_2010a,zhangShu_2010b}, we develop a realizability-preserving method to simulate transport of particles (fermions) through a background material using the two-moment model of particle transport, which evolves angular moments of a phase space distribution function $f$.  
The two-moment model is closed using algebraic moment closures; e.g., as proposed by Cernohorsky \& Bludman \cite{cernohorskyBludman_1994} and Banach \& Larecki \cite{banachLarecki_2017a}.  
Variations of this model have recently been used to simulate neutrino transport in nuclear astrophysics applications --- including core-collapse supernovae and compact binary mergers.  
We employ the discontinuous Galerkin (DG) method for spatial discretization (in part to capture the asymptotic diffusion limit of the model) combined with implicit-explicit (IMEX) time integration to stably bypass short timescales induced by frequent interactions between particles and the background.  
Appropriate care is taken to ensure the method preserves strict algebraic bounds on the evolved moments (particle density and flux) as dictated by Pauli's exclusion principle, which demands a bounded distribution function (i.e., $f\in[0,1]$).  
This realizability-preserving scheme combines a suitable CFL condition, a realizability-enforcing limiter, a closure procedure based on Fermi-Dirac statistics, and an IMEX scheme whose stages can be written as a convex combination of forward Euler steps combined with a backward Euler step.  
The IMEX scheme is formally only first-order accurate, but works well in the diffusion limit, and --- without interactions with the background --- reduces to the optimal second-order strong stability-preserving explicit Runge-Kutta scheme of Shu \& Osher \cite{shuOsher_1988}.  
Numerical results demonstrate the realizability-preserving properties of the scheme.  
We also demonstrate that the use of algebraic moment closures not based on Fermi-Dirac statistics can lead to unphysical moments in the context of fermion transport.  
\end{abstract}

>>>>

%% Group authors per affiliation:
\author[utk-phys]{Ran Chu}
\ead{rchu@vols.utk.edu}

\author[ornl,utk-phys,jics]{Eirik Endeve\corref{cor}}
\ead{endevee@ornl.gov}

@@

\begin{abstract}
Building on the framework of Zhang \& Shu \cite{zhangShu_2010a,zhangShu_2010b}, we develop a realizability-preserving method to simulate the transport of particles (fermions) through a background material using a two-moment model that evolves the angular moments of a phase space distribution function $f$.  
The two-moment model is closed using algebraic moment closures; e.g., as proposed by Cernohorsky \& Bludman \cite{cernohorskyBludman_1994} and Banach \& Larecki \cite{banachLarecki_2017a}.  
Variations of this model have recently been used to simulate neutrino transport in nuclear astrophysics applications, including core-collapse supernovae and compact binary mergers.  
We employ the discontinuous Galerkin (DG) method for spatial discretization (in part to capture the asymptotic diffusion limit of the model) combined with implicit-explicit (IMEX) time integration to stably bypass short timescales induced by frequent interactions between particles and the background.  
Appropriate care is taken to ensure the method preserves strict algebraic bounds on the evolved moments (particle density and flux) as dictated by Pauli's exclusion principle, which demands a bounded distribution function (i.e., $f\in[0,1]$).  
This realizability-preserving scheme combines a suitable CFL condition, a realizability-enforcing limiter, a closure procedure based on Fermi-Dirac statistics, and an IMEX scheme whose stages can be written as a convex combination of forward Euler steps combined with a backward Euler step.  
The IMEX scheme is formally only first-order accurate, but works well in the diffusion limit, and --- without interactions with the background --- reduces to the optimal second-order strong stability-preserving explicit Runge-Kutta scheme of Shu \& Osher \cite{shuOsher_1988}.  
Numerical results demonstrate the realizability-preserving properties of the scheme.  
We also demonstrate that the use of algebraic moment closures not based on Fermi-Dirac statistics can lead to unphysical moments in the context of fermion transport.  
\end{abstract}
